### 2023-03-27 main/chapter_searches.tex
add heplikelihood formula

Despite their troublesome nomenclature, \heplikelihood\ models are extremely
useful and effective for describing the predictions of our hypotheses and
extracting interpretable results.
A general \heplikelihood\ model is a function $H(x)$ that takes input
parameters $x$ and outputs two things:
expected yields $y(x)$ to predict Poisson data,
and an objective value $E$ to be minimized (in a `fit') to choose a best
prediction.
That is,
\begin{equation}
H(x) = [\,y(x), \,E(x)\,]
.
\end{equation}
The minimization objective $E$ is the summed negative $\log$ \heplikelihood\
for all actual and auxiliary data.
We choose the letter $E$ for \textbf{energy}, to highlight the analogy with
statistical physics and the Boltzmann factor
$f(x) \propto e^{-E(x)/(k_B T)}$,
with which a thermodynamic system collapses to its minimum energy at cold
temperatures~\cite{
skilling2017david,
lecun-06,
pmlr-v2-ranzato07a
}.
At unit temperature, this can induce a probability distribution
relative to an appropriate measure~\cite{
cranmer2012histfactory,
skilling2017david
},
but in my experience with \heplikelihood\ models such distributions often
coincide with neither intuition nor the best fit value.%
\footnote{%
One can reproduce this experience by sampling a public \pyhf\ \heplikelihood\
model
(for example, try SR-HM(disc.) of $1\ell bb$~\cite{SUSY-2019-08,hepdata.90607})
with standard Markov-Chain Monte Carlo algorithms.
I recommend first estimating the covariance by expanding about the maximum
\heplikelihood, and linearly transforming your coordinates to whiten that
covariance to the identity matrix.
Then, algorithms provided by
TensorFlow Probability~\cite{tensorflow2015-whitepaper}
(the Metropolis-Adjusted Langevin Algorithm~\cite{mala1998, XIFARA201414}
is a good start)
can sample quite efficiently with tensor evaluation through
\pyhf~\cite{heinrich2021pyhf}.
Histograms of samples for the signal strength in the discovery fits
can differ substantially from profile-\heplikelihood\ curves.
}

>>>>

Despite their troublesome nomenclature, \heplikelihood\ models are extremely
useful and effective for describing the predictions of our hypotheses and
extracting interpretable results.
A general \heplikelihood\ model is a function $H(x)$ that takes input
parameters $x$ and outputs two things:
expected yields $y(x)$ to predict Poisson data,
and an objective value $E$ to be minimized (in a `fit') to choose a best
prediction.
That is,
\begin{equation}
H(x) = [\,y(x), \,E(x)\,]
.
\end{equation}
The minimization objective $E$ is the summed negative $\log$ \heplikelihood\
for all actual and auxiliary data.
We choose the letter $E$ for \textbf{energy}, to highlight the analogy with
statistical physics and the Boltzmann factor
$f(x) \propto e^{-E(x)/(k_B T)}$,
with which a thermodynamic system collapses to its minimum energy at cold
temperatures~\cite{
skilling2017david,
lecun-06,
pmlr-v2-ranzato07a
}.
At temperature $1/k_B$, this can induce a probability distribution
relative to an appropriate measure~\cite{
cranmer2012histfactory,
skilling2017david
},
but in my experience with \heplikelihood\ models such distributions often
coincide with neither intuition nor the best fit value.%
\footnote{%
One can reproduce this experience by sampling a public \pyhf\ \heplikelihood\
model
(for example, try SR-HM(disc.) of $1\ell bb$~\cite{SUSY-2019-08,hepdata.90607})
with standard Markov-Chain Monte Carlo algorithms.
I recommend first estimating the covariance by expanding about the maximum
\heplikelihood, and linearly transforming your coordinates to whiten that
covariance to the identity matrix.
Then, algorithms provided by
TensorFlow Probability~\cite{tensorflow2015-whitepaper}
(the Metropolis-Adjusted Langevin Algorithm~\cite{mala1998, XIFARA201414}
is a good start)
can sample quite efficiently with tensor evaluation through
\pyhf~\cite{heinrich2021pyhf}.
Histograms of samples for the signal strength in the discovery fits
can differ substantially from profile-\heplikelihood\ curves.
}

The formula for a standard \roofit\ \heplikelihood\ function can be
summarised as
\begin{equation}
\label{eqn:searches_heplikelihood_e}
e^{E(x)} =
\prod_{b \in \mathrm{bins}}\mathrm{Poisson}(n_b \mid y_b(x))
\times \prod_{i \in \mathrm{parameters}} f_i(x_i),
\end{equation}
in which $\mathrm{Poisson}(n\mid \mu)$ is the Poisson likelihood function
of Equation~\ref{eqn:searches_poisson}, $y_b(x)$ is the yield in bin $b$
for parameters $x$, and $f_i$ is the constraint function for parameter $i$.
Constraint functions $f_i$ are typically of
Gaussian (Equation~\ref{eqn:searches_gaussian}) or
Poissonian (Equation~\ref{eqn:searches_poisson} or a continuous interpolation)
forms, and often interpreted as likelihoods on fictional data.
The mappings from parameters to yields $y_b(x)$ are parameterised to
non-linearly interpolate between given `low', `central', and `high' variations
of the yields corresponding to certain parameters; many such
interpolation/extrapolation functions are available, but the most common
interpolates with a sixth-order polynomial that is smoothly attached to
exponential tails for extrapolation.
To avoid the numerical difficulties associated with taking a product over many
terms, one typically handles $E(x)$, not $e^{E(x)}$.
The bins may be further grouped into channels that share
terms~\cite{verkerke2003roofit}.
